---
title: "Personal Finance Analysis using R"
author: "Miguel Castillo"
date: "03/01/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Data is everywhere and we can make good use of it by starting with our personal expenses.
The data being used here was created just for educational purposes to explain how the R Script created works.
You can always change the data set used with one with your own expenses.

### Libraries

The R script uses these libraries:
* Lubridate
* tidiverse

```{r include=FALSE, message=FALSE, warning=FALSE}
library(lubridate)
library(tidyverse)
library(dplyr)
```

## The Data Set

The data set was created as a CSV file using a Spreadsheet application and it consist of 7 columns and 42 rows.

```{r include=FALSE, message=FALSE, warning=FALSE}
# Loading the data set for expenses
expenses_dataset = read.csv(file = 'expenses_dataset.csv')
```

One of the 7 columns is called date_of_purchased, which is then separated into days, months and years for subsequent processing.
The month column is then converted to text format for reading easiness.


```{r message=FALSE, warning=FALSE}
expenses_dataset <- separate(expenses_dataset, date_of_purchase, into=c('day','month','year'),sep='/')
months_name <- c('Jan', 'Feb', 'Mar', 'Apr', 'May', 'Jun', 'Jul', 'Aug', 'Sep', 'Oct', 'Nov', 'Dec')
expenses_dataset$day <- as.integer(expenses_dataset$day)
expenses_dataset$month <- as.integer(expenses_dataset$month)
expenses_dataset$year <- as.integer(expenses_dataset$year)
expenses_dataset$price <- as.numeric(expenses_dataset$price)
```

This is a preview of the data set.

```{r warning=FALSE}
head(expenses_dataset)
```

```{r include=FALSE, message=FALSE, warning=FALSE}
# The initial year from the data set
minyear <- min(expenses_dataset$year)
# The last year from the data set
maxyear <- max(expenses_dataset$year)
```

## Rearreanging method

The data set will be rearranged in to 3 different data sets for specific analysis purposes.

Creating the category_expense data frame.
In this way, we can group the expenses by category and getting the total for each category, finally, sorting them in descending order.

```{r warning=FALSE}
category_expense <- expenses_dataset %>%
  group_by(category) %>%
  summarise(total = sum(price)) %>%
  arrange(desc(total))
```

Then, creating the month_expense data frame.
In this way, we can group by month and getting the total expense for each month.

```{r warning=FALSE}
month_expense <- expenses_dataset %>%
  group_by(month) %>%
  summarise(total = sum(price))
# Converting months to name for easiness reading
month_expense$month <- months_name[as.integer(month_expense$month)]
# Creating vector of months in numeric value to get the trend line
numeric_months <- match(month_expense$month,month.abb)
```

Finally, Creating the year_expense data frame.
In this way, we can group by year and getting the total amount spend per year.
```{r warning=FALSE}
year_expense <- expenses_dataset %>%
  group_by(year) %>%
  summarise(total = sum(price))
```

## Plotting and analyzing


Plotting the category_expense data set
Using columns (bars) and ordering in descending total order using reorder()
Filling each bar from green to red to visually see in which category did we spent the most.

```{r warning=FALSE}
ggplot(category_expense) + 
  geom_col(mapping = aes(x=reorder(category, -total), y = total, fill = total)) + 
  scale_fill_gradient(low = "green", high = "red", space = 'Lab') + 
  theme(axis.text.x = element_text(angle = 90, size = 11), axis.text.y = element_text(angle = 90, size = 11), axis.title.x = element_text(size = 15), axis.title.y = element_text(size = 15)) +
  labs(title = 'Expenses grouped by Category', subtitle = paste0("Data from: ", minyear, " to: ", maxyear), x = "Categories", y = 'Total (MXN)')
```
As seen, the expenses related to the House are the ones that reached a high amount of money spent compared to the next category (Lodging).

By using the filter() function we can filter by Category == "House" and see why the expenses are far higher than the others.

```{r warning=FALSE}
filter(expenses_dataset, category == "House")
```

Then, get the plot from the month_expense data set.
Again, using columns (bars) and filling each bar from green to red.

```{r warning=FALSE}
ggplot(month_expense) + 
  geom_col(mapping = aes(x=fct_inorder(month), y=total, fill=total)) + 
  scale_fill_gradient(low="green", high="red", space='Lab') + 
  geom_smooth(aes(x=numeric_months, y=total), method=loess, color="blue", formula=y ~ x, se=FALSE ,linetype = "dashed") + 
  theme(axis.text.x = element_text(angle = 90, size = 11), axis.text.y = element_text(angle = 90, size = 11), axis.title.x = element_text(size = 15), axis.title.y = element_text(size = 15)) +
  labs(title = 'Expenses grouped by Month', subtitle = paste0("Data from: ", minyear, " to: ", maxyear, " with trend line using y=x and using the loesses method"), x = "Month", y = 'Total (MXN)')
```

It seems that the months of July and September are the ones with a higher expense rate, could be because of the vacation season or any other reason that requires to be reviewed within the data set, so it can be confirmed if the expenses were caused due to a 1-time event or something that can be decreased by taking specific actions.

Finally, plotting the year_expense data set.
Using columns (bars) and filling each bar from green to red.

```{r warning=FALSE}
ggplot(year_expense) + geom_col(mapping = aes(x=year, y=total, fill=total)) + 
  scale_fill_gradient(low="green", high="red", space='Lab') + 
  geom_smooth(aes(x=year, y=total), method=glm, color="blue", formula=y ~ x, se=FALSE, linetype = "dashed") + 
  theme(axis.text.x = element_text(angle = 90, size = 11), axis.text.y = element_text(angle = 90, size = 11), axis.title.x = element_text(size = 15), axis.title.y = element_text(size = 15)) +
  labs(title = 'Expenses grouped by Year', subtitle = paste0("Data from: ", minyear, " to: ", maxyear, " with trend line using y=x and using the Generalized linear model"), x = "Year", y = 'Total (MXN)')
```
The trendline shows a excrement of money spent from 2018 to 2020, which predicts a decrease of expenses for 2021 and 2022.
Of course, this data set only contains information from 3 years and the conclusions that can be generated from this information can be biased or inaccurate, and more data should require to be gathered.

## Conclusion

This R Script can help you have an understanding of your expenses and make some decisions to improve them based on available and raw data.



